The low dimensional world of echolocating bats

### 2020-03-13 Paper/main.tex
before running paper through grammarly

One mechanism for place recognition would be for bats to parse the stream of echoes returning from the environment \cite[e.g.,][]{Yovel2009,Vanderelst2016} into localized and identified objects. However, as we argued before, it is highly questionable whether bats can interpret echoes in terms of a 3D environment model or acoustic image \citep[e.g.,][]{Vanderelst2015,Vanderelst2016,Steckel2013}. Sonar maps the three-dimensional structure of the environment on two one-dimensional waveforms arriving at the ears. While the time of arrival of echoes encodes object distance, azimuth and elevation directions are not directly available from the acoustic signal. In principle, azimuth and elevation can be estimated using spectral cues imposed by the head-related transfer function. However, complex environments return many overlapping and interfering echoes, making reconstructing the direction of origin of echoes much harder, and perhaps impossible. Extracting spatial information is further hampered by the limited temporal resolution of the sonar system \citep{Simmons1989,Wiegrebe1996,Surlykke1996}.

We have suggested before that limitations of the bat's sonar system result in bats having access to a sparse, low dimensional description of the outside 3D world only \citep{Vanderelst2015a,Vanderelst2016}. Some recent evidence seems to confirm this. First, \citet{Geberl2019} confirmed that bio-sonar has a very low spatial resolution, both for distance and angular dimensions. In addition, neurophysiological data suggest that closely spaced objects are not perceived individually by echolocating bats. In contrast, neural responses to echo cascades from densely spaced objects seem to represent a single extended stimulus \citep{Warnecke2018}. \todo[inline]{Look at recent Simmons paper where bats refuse to fly through hoop}

@@

In this paper, we directly evaluate the hypothesis that sonar provides bats inherently with a low dimensional, yet sufficiently rich, description of the world.

We test this hypothesis in two steps. In a first step, we generate a large collection of synthetic echo sequences. We assess the compressibility of these echo signals using \PCA. Next, we filter the echo sequences using a model of the bat's auditory periphery \citep{Wiegrebe2008}. The synthetic echo sequences differ greatly in complexity. Nevertheless, we show that they can be compressed considerably. Moreover, the output of the cochlear model can be mapped onto an even lower dimensional space. In a second step, we show that echoes collected in real bat habitats can be mapped into the same lower dimensional space, without loss of information. We do this by demonstrating that the low dimensional feature space allows recognizing locations in real habitats.

Together these results suggest that echolocating bats live in a low dimensional perceptual world. The physics of the sonar system and the auditory periphery map the complexity of the 3D environment onto a small number of features, irrespective of whether we use artificial or real echo sequences. Nevertheless, these features are sufficiently rich for bats to recognize locations, and as we discuss, support flight control.

>>>>

One mechanism for place recognition would be for bats to parse the stream of echoes returning from the environment into localized and identified objects [REF]. However, as we argued before, it is highly questionable whether bats can interpret echoes in terms of a 3D environment model or acoustic image \citep[e.g.,][]{Vanderelst2015,Vanderelst2016,Steckel2013}. Sonar maps the three-dimensional structure of the environment on two one-dimensional waveforms arriving at the ears. While the time of arrival of echoes encodes object distance, azimuth and elevation directions are not directly available from the acoustic signal. In principle, azimuth and elevation can be estimated using spectral cues imposed by the head-related transfer function [REF]. However, complex environments return many overlapping and interfering echoes, making reconstructing the direction of origin of echoes much harder, and perhaps impossible. Extracting spatial information is further hampered by the limited temporal resolution of the sonar system \citep{Simmons1989,Wiegrebe1996,Surlykke1996}.

We have suggested before that limitations of the bat's sonar system result in bats having access to a sparse, low dimensional description of the outside 3D world only \citep{Vanderelst2015a,Vanderelst2016}. Some recent evidence seems to confirm this. First, \citet{Geberl2019} confirmed that bio-sonar has a low spatial resolution, both for distance and angular dimensions. Also, neurophysiological data suggest that closely spaced objects are not perceived seperatly by echolocating bats. In contrast, neural responses to echo cascades from densely spaced objects seem to represent a single extended stimulus \citep{Warnecke2018}. \todo[inline]{Look at recent Simmons paper where bats refuse to fly through hoop}

@@

In this paper, we directly evaluate the hypothesis that sonar provides bats inherently with a low dimensional, yet sufficiently detailed, description of the world. We test this hypothesis in two steps. In a first step, we generate an extensive collection of synthetic echo sequences. We assess the compressibility of these echo signals using \PCA. Next, we filter the echo sequences using a model of the bat's auditory periphery \citep{Wiegrebe2008}. The synthetic echo sequences differ greatly in complexity. Nevertheless, we show that they can be compressed considerably. Moreover, the output of the cochlear model can be mapped onto an even lower-dimensional space. In a second step, we show that echoes collected in real bat habitats can be mapped into the same low-dimensional space, without loss of information. We do this by demonstrating that the low dimensional feature space allows recognizing locations in real habitats.

Together these results suggest that echolocating bats live in a low dimensional perceptual world. The physics of the sonar system and the auditory periphery map the complexity of the 3D environment onto a small number of independent features, irrespective of whether we use artificial or real echo sequences. Nevertheless, these features are sufficiently rich for bats to recognize locations, and as we discuss, support flight control.
